---
title: "Introduction to Social Networks"
author: "Matteo Larrode"
date: "2024-01-11"
output: html_document
---

### Exercise 1 - 3

```{r}
library(igraph)
library(network)
library(igraphdata)

data(karate)
```

```{r}
print(karate)
```
In this output:
- first line = title
- attributes
  - 'g' means graph = general information about the network
  - 'v' means vertex = information about vertices
  - 'e' means edge = information about edges
  - 'c' means character (class)
  - 'n' means numeric (class)
- list of edges

```{r}
plot(karate)
```
### Exercise 4

```{r}
data(flo)
print(flo)
```
This data is loaded not as an 'igraph' object, but as an adjacency matrix! Hence the printing looks different. 

To plot it, we need to turn it into a network object:

```{r}
g <- graph_from_adjacency_matrix(flo, mode = "undirected")

plot(g)
```
### Exercise 5

```{r}
D <- read.csv("data/practical1.csv", header = TRUE)

head(D)
```
```{r}
D$tie <- ifelse(D$Emails<5, 0, 1) 
D2 <- subset(D, Emails >=5)
D2 <- subset(D2, select = -Emails) 

write.csv(D2, file = "data/netbinary.csv", row.names = FALSE) 

```

Can you imagine the new dataset you created in 5-b (netbinary) as a 5-by-5 matrix?
Note, think about the ``diagonals’’ of this matrix, what could they be?

The datasets we have worked with before (D & D2) are edge lists, now we will be creating an adjacency matrix. 

```{r}
M <- matrix(c(0, 1, 1, 1, 0,
              1, 0, 0, 0, 1,
              0, 1, 0, 1, 0,
              1, 0, 1, 0, 1, 
              0, 1, 0, 1, 0), nrow = 5, ncol = 5, byrow = TRUE)

rownames(M) <- c("Ozan", "Burak", "Matteo", "Nad", "Zoe")
colnames(M) <- c("Ozan", "Burak", "Matteo", "Nad", "Zoe")

print(M)


```
We can also plot it:

```{r}
g2 <- graph_from_adjacency_matrix(M) 

plot(g2) 
```